Asteroseismology: AI-Driven Seismic Event Detection for Extra-Terrestrial Missions\\

### 2025-05-25 paper.tex
update, potential checkpoint

        Decades later, the NASA InSight mission, which landed in 2018, successfully deployed the SEIS (Seismic
        Experiment for Interior Structure) instrument directly onto the Martian surface \cite{Lognonne2019, Banerdt2020}. 
        %(Lognonné et al., 2019; Banerdt et al., 2020) 
        SEIS comprises a suite of highly sensitive three-component Very Broad Band (VBB) and three-component Short
        Period (SP) seismometers. InSight has definitively recorded hundreds of marsquakes, confirming Mars as a
        seismically active planet \cite{Giardini2020,Clinton2021}. Martian seismic signals, like lunar ones, can be
        influenced by near-surface scattering, though generally less intense than on the Moon. The Marsquake Service
        (MQS) is responsible for ongoing event detection and cataloging, employing a combination of traditional and
        evolving ML-assisted techniques \cite{Clinton2021}.

@@

        \subsubsection{STA\slash LTA Analysis for Initial Detection}
            Potential seismic event onsets are identified using the classic Short-Term Average to Long-Term Average
            (STA/LTA) algorithm, as implemented in Obspy \cite{Beyreuther2010}. The algorithm is applied to the absolute
            values of the normalized and filtered trace data. The lengths of the short-term window and long-term window,
            as well as the trigger-on threshold and trigger-off threshold, are configurable parameters. The trigger-off
            threshold is dynamically set based on the average of the characteristic function (CFT) generated by the
            STA/LTA algorithm. This results in a set of on-off time pairs, marking the start and end times of candidate
            detections.An example of the filtered seismogram alongside its STA/LTA characteristic function, with trigger
            points indicated, is shown in ``Fig.~\ref{fig:stalta_output}''.
            \begin{figure}[htbp]
                \centerline{\includegraphics[width=0.9\columnwidth]{figures/fig4_stalta_detections.pdf}}
                \caption{Output of the STA/LTA analysis. The top panel shows the filtered and normalized seismogram. 
                The bottom panel displays the corresponding STA/LTA characteristic function, with red vertical 
                lines indicating trigger-on times for detected candidate events.}
                \label{fig:stalta_output}
            \end{figure}

@@

Other studies have explored ML for analyzing data from the InSight mission to Mars, focusing on tasks like marsquake
detection and noise characterization \cite{b21}. The noisy operational environment of landers, combined with the diverse
and often weak seismic signals, makes robust event detection particularly challenging and an ideal candidate for
ML-driven solutions.

Our Asteroseismology framework builds upon these foundations. It acknowledges the strengths of conventional algorithms
like STA/LTA for initial, computationally cheap candidate generation, similar to some multi-stage earthquake detection
systems on Earth. However, it significantly differs from approaches like Civilini et al. \cite{favPaper} by using a
hybrid pipeline that feeds 1D waveform segments and engineered auxiliary features directly to the CNN, rather than
relying solely on spectrograms. This aims to retain fine temporal details and leverage interpretable features,
potentially enhancing both efficiency and robustness for onboard processing.

@@

% In conclusion, the framework represents a significant step towards enabling more autonomous and scientifically
% productive planetary seismology missions. By intelligently managing data at the source, it helps to overcome critical
% mission constraints, paving the way for deeper and more comprehensive exploration of the internal structures of
% celestial bodies in our solar system.
\section{Conclusion and Future Work}
This paper has presented Asteroseismology, an AI-driven hybrid framework designed for efficient seismic event detection
on extra-terrestrial missions. Our system combines classical seismic processing with a lightweight Convolutional Neural
Network (CNN) and demonstrates strong potential for autonomous, onboard data analysis. The framework's rapid processing
capability (a month of data in under 60 seconds), adaptability through tunable parameters, and initial detection
accuracy exceeding 80\% with a false positive rate around 5%, highlight its suitability for optimizing data telemetry
from missions to the Moon, Mars, and potentially other planetary bodies.

@@

To further enhance the Asteroseismology framework, future efforts can be directed towards several key areas:
\begin{enumerate}
\item \textit{Advanced CNN Development and Training Data Expansion:} Future iterations will explore more sophisticated
CNN architectures. Expanding the training dataset with a greater variety of seismic event types and noise profiles,
along with advanced data augmentation techniques, will be crucial for improving the model's generalization and
robustness across different planetary conditions.
\item \textit{Generative Adversarial Networks (GANs) for Data Augmentation and Feature Learning:} Inspired by
applications in terrestrial seismology where GANs have been used for tasks like seismic wave discrimination (e.g., \cite{Li2018} who used a GAN critic as a feature extractor), we plan to investigate the use of GANs, potentially AC-GANs
\cite{Odena2017}, to generate realistic synthetic planetary seismic data. This could be particularly beneficial for
augmenting training sets with rare event types and improving the classifier's ability to distinguish subtle event
characteristics.
\item \textit{Physics-Informed Neural Networks (PINNs) and Enhanced Auxiliary Features:} We will investigate the
inclusion of a richer set of auxiliary features for the CNN, derived from more detailed physical analysis of candidate
event waveforms. Furthermore, exploring Physics-Informed Neural Networks (PINNs), which are gaining momentum in
seismology (Mousavi and Beroza, 2022), could allow for the incorporation of known physical constraints directly into the
ML model, potentially leading to more robust and interpretable event detection.
\item \textit{Onboard Implementation and Prototyping:} A critical next step is to prototype and rigorously evaluate the
Asteroseismology system on hardware representative of spacecraft processors (e.g., FPGAs or radiation-hardened CPUs) to
assess its performance under realistic mission resource constraints.
\item \textit{Adaptation to Other Celestial Bodies:} The framework's adaptable design makes it a promising candidate for
future missions to other seismically active bodies, such as Europa or Titan. This would involve tuning parameters and
retraining the CNN with appropriate simulated or acquired data specific to those environments.
\end{enumerate}
In conclusion, the Asteroseismology framework offers a significant advancement towards more autonomous and scientifically productive planetary seismology missions. By enabling intelligent data management at the source, it helps overcome critical mission constraints, paving the way for deeper exploration and understanding of the internal structures of celestial bodies within our solar system.

\section*{Acknowledgment}
The authors would like to thank the NASA Planetary Data System (PDS) for providing access to the Apollo ALSEP and Mars
InSight mission data. We also acknowledge the developers of open-source software packages used in this research,
including.%TODO; acknolees

@@

    % \bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350. 

@@

    % \bibitem{b22} A. Odena, C. Olah, and J. Shlens, ``Conditional image synthesis with auxiliary classifier GANs,'' in \emph{Proc. Int. Conf. Mach. Learn. (ICML)}, 2017, pp. 2642--2651.

>>>>

        Decades later, the NASA InSight mission, which landed in 2018, successfully deployed the SEIS (Seismic
        Experiment for Interior Structure) instrument directly onto the Martian surface \cite{Lognonne2019, Banerdt2020}. 
        SEIS comprises a suite of highly sensitive three-component Very Broad Band (VBB) and three-component Short
        Period (SP) seismometers. InSight has definitively recorded hundreds of marsquakes, confirming Mars as a
        seismically active planet \cite{Giardini2020,Clinton2021}. Martian seismic signals, like lunar ones, can be
        influenced by near-surface scattering, though generally less intense than on the Moon. The Marsquake Service
        (MQS) is responsible for ongoing event detection and cataloging, employing a combination of traditional and
        evolving ML-assisted techniques \cite{Clinton2021}.

@@

        \subsubsection{STA\slash LTA Analysis for Initial Detection}
            Potential seismic event onsets are identified using the classic Short-Term Average to Long-Term Average
            (STA/LTA) algorithm, as implemented in Obspy \cite{Beyreuther2010}. The algorithm is applied to the absolute
            values of the normalized and filtered trace data. The lengths of the short-term window and long-term window,
            as well as the trigger-on threshold and trigger-off threshold, are configurable parameters. The trigger-off
            threshold is dynamically set based on the average of the characteristic function (CFT) generated by the
            STA/LTA algorithm. This results in a set of on-off time pairs, marking the start and end times of candidate
            detections.An example of the filtered seismogram alongside its STA/LTA characteristic function, with trigger
            points indicated, is shown in ``Fig.~\ref{fig:stalta_output}''.
            \begin{figure}[htbp]
                \centerline{\includegraphics[width=0.9\columnwidth]{figures/fig4_stalta_detections.pdf}}
                \caption{Output of the STA/LTA analysis. (a) shows the filtered and normalized seismogram. 
                (b) displays the corresponding STA/LTA characteristic function, with red horizontal 
                line indicating trigger-on threshold for detected candidate events.}
                \label{fig:stalta_output}
            \end{figure}

@@

% Other studies have explored ML for analyzing data from the InSight mission to Mars, focusing on tasks like marsquake
% detection and noise characterization \cite{b21}. The noisy operational environment of landers, combined with the diverse
% and often weak seismic signals, makes robust event detection particularly challenging and an ideal candidate for
% ML-driven solutions.

% Our Asteroseismology framework builds upon these foundations. It acknowledges the strengths of conventional algorithms
% like STA/LTA for initial, computationally cheap candidate generation, similar to some multi-stage earthquake detection
% systems on Earth. However, it significantly differs from approaches like Civilini et al. \cite{favPaper} by using a
% hybrid pipeline that feeds 1D waveform segments and engineered auxiliary features directly to the CNN, rather than
% relying solely on spectrograms. This aims to retain fine temporal details and leverage interpretable features,
% potentially enhancing both efficiency and robustness for onboard processing.

@@

% In conclusion, the framework represents a significant step towards enabling more autonomous and scientifically
% productive planetary seismology missions. By intelligently managing data at the source, it helps to overcome critical
% mission constraints, paving the way for deeper and more comprehensive exploration of the internal structures of
% celestial bodies in our solar system.
\section{Conclusion and Future Work}
This paper has presented Asteroseismology, an AI-driven hybrid framework designed for efficient seismic event detection
on extra-terrestrial missions. Our system combines classical seismic processing with a Convolutional Neural
Network (CNN) and demonstrates strong potential for autonomous, onboard data analysis. The framework's rapid processing
capability (a month of data in under 60 seconds), adaptability through tunable parameters, and initial detection
accuracy exceeding 80\% with a false positive rate around 5\%, highlight its suitability for optimizing data telemetry
from missions to the Moon, Mars, and potentially other planetary bodies.

@@

To further enhance the Asteroseismology framework, future efforts can be directed towards several key areas:
\begin{enumerate}
\item \textit{Advanced CNN Development and Training Data Expansion:} Future iterations can explore more sophisticated
CNN architectures. Expanding the training dataset with a greater variety of seismic event types and noise profiles,
along with advanced data augmentation techniques, will be crucial for improving the model's generalization and
robustness across different planetary conditions.
\item \textit{Generative Adversarial Networks (GANs) for Data Augmentation and Feature Learning:} Inspired by
applications in terrestrial seismology where GANs have been used for tasks like seismic wave discrimination (e.g.,
\cite{Li2018} who used a GAN critic as a feature extractor), we can investigate the use of GANs, potentially AC-GANs
\cite{Odena2017}, to generate realistic synthetic planetary seismic data. This could be particularly beneficial for
augmenting training sets with rare event types and improving the classifier's ability to distinguish subtle event
characteristics.
\item \textit{Physics-Informed Neural Networks (PINNs) and Enhanced Auxiliary Features:} We will investigate the
inclusion of a richer set of auxiliary features for the CNN, derived from more detailed physical analysis of candidate
event waveforms. Furthermore, exploring Physics-Informed Neural Networks (PINNs), which are gaining momentum in
seismology \cite{Mousavi2022}, could allow for the incorporation of known physical constraints directly into the
ML model, potentially leading to more robust and interpretable event detection.
\item \textit{Onboard Implementation and Prototyping:} A critical next step is to prototype and rigorously evaluate the
system on hardware representative of spacecraft processors (e.g., FPGAs or radiation-hardened CPUs) to
assess its performance under realistic mission resource constraints.
\item \textit{Adaptation to Other Celestial Bodies:} The framework's adaptable design makes it a promising candidate for
future missions to other seismically active bodies, such as Europa or Titan. This would involve tuning parameters and
retraining the CNN with appropriate simulated or acquired data specific to those environments.
\end{enumerate}
In conclusion, the framework offers an advancement towards more autonomous and scientifically productive planetary
seismology missions. By enabling intelligent data management at the source, it helps overcome critical mission
constraints, paving the way for deeper exploration and understanding of the internal structures of celestial bodies
within our solar system.

\section*{Acknowledgment}
The authors would like to express their sincere gratitude to the NASA Planetary Data System (PDS), including its
Geosciences Node and the IRIS Data Management Center (IRIS-DMC), for providing public access to the invaluable seismic
data from the Apollo Lunar Surface Experiments Package (ALSEP) missions. We also thank NASA, CNES, and all international
partners of the Mars InSight mission for making the SEIS instrument data available through the PDS and services like
IRIS-DMC. These open data policies are fundamental to advancing planetary seismology. We also acknowledge the crucial
contributions of the open-source software community. This research made extensive use of several packages, including:
\begin{itemize}
\item \textit{Obspy} \cite{Beyreuther2010} for seismic data processing and analysis;
\item \textit{Keras} \cite{Chollet2021} with \textit{TensorFlow} \cite{Abadi2016} as its backend for developing and
training the Convolutional Neural Network;
\end{itemize}
The availability of these powerful tools significantly accelerates research and development in this field.

@@

    \bibitem{Odena2017} A. Odena, C. Olah, and J. Shlens, ``Conditional image synthesis with auxiliary classifier GANs,'' in \emph{Proc. Int. Conf. Mach. Learn. (ICML)}, 2017, pp. 2642--2651.
    \bibitem{Mousavi2022} S. M. Mousavi and G. C. Beroza, ``Deep-learning seismology,'' \emph{Science}, vol. 377, no. 6607, p. eabm4470, Aug. 2022.   % \bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350. 
    \bibitem{Abadi2016} M. Abadi et al., "TensorFlow: A System for Large-Scale Machine Learning," in Proc. 12th USENIX Symp. Operating Syst. Design Implementation (OSDI), 2016, pp. 265--283.

@@
